Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-07-09 main.tex

\noindent
In order to compute the above function $b_\bot : Q \times Q \to \Semiring$, 
we shall consider an auxiliary function $b_\top :  Q \times P \times Q \to \Phic$.
%
Intuitively, when $b_\top(q, p, q') = \phi \in \Phic$, 
for all $c \in \Deltac$,  $\phi(c)$ is 
the minimum weight of a computation of~$A$
starting in state $q$ with a non-empty stack 
$\gamma' = \<c, p> \, \gamma \in \Gamma^+$ 
and ending in state $q'$ with the same stack $\gamma'$,
such that moreover, 
the computation does not pop the pair $\<c, p>$ at the top of $\gamma'$
(\ie $\gamma'$ is left untouched during the computation).
However, the computation can read $\<c, p>$ at the top of $\gamma'$,
and can also push another pair $\< c', p'> \in \Gamma$ on $\gamma'$,
following the third case of 
in the definition \eqref{eq:SWVPA-weight} of $\weight_A$ (call symbol).
The pair $\< c', p'>$ can be pop later, during the computation from $q$ to $q'$,  
following the fifth case of \eqref{eq:SWVPA-weight} (return symbol). 
%However, it cannot apply one of the two last cases (return symbol and empty stack)
%when the current stack is $\gamma$.
%pop symbols in $\gamma$.
%
Formally, in order to define $b_\top$, we consider 
a fresh stack symbol $\top \notin \Gamma$,   %which does not belong to $\Gamma$,
representing the above untouched stack, and let:
%
\begin{equation}\label{eq:btop}
  b_\top(q, p, q') : c \mapsto \bigoplus_{s\in \Delta^*}
  \textstyle
  \weight_A\bigl(\configup{q}{\< c, p> \stackup \top }, s, \configup{q'}{\<c, p> \stackup \top}\bigr)
\quad  
\mathrm{~for~all~} c \in \Deltac  
\end{equation}
%
By definition of $\weight_A$ in \eqref{eq:SWVPA-weight}, 
using the symbol $\top$ for the part of the stack below $\<c, p>$
(\ie the substack $\gamma$ in the above $\gamma' = \<c, p> \, \gamma$)
ensures that this part $\gamma'$ is not touched during the computation.
% there cannot be a pop of $\<c, p>$ followed by a push of $\<c, p>$
% because it is not possible to push over $\top$ (push read the top of stack).
%
This ensures in particular that the subword 
read during the computation is well parenthesized
(every symbol in $\Deltac$ has a matching symbol in $\Deltar$).

@@

\item[--] a \SWVPA $A$ over $\Delta$, $\Semiring$, and $\bar\Phi$,
defining a measure $A: \Delta^* \to \Semiring$,
%a \SWVPA $A$ over $\Delta$, and $\Semiring$, defining a series of nested words
%      $A : \Delta^* \to \Semiring$,
\item[--] an input word $s \in \Sigma^*$.
\end{description}
%
For all $u \in \Sigma^*$ and $t \in \Delta^*$, let
\(d(u, t) = T\bigl(u, t|_\Deltai \bigr)\),
where $t|_\Deltai \in \Deltai^*$ is the projection of $t$ onto~$\Deltai$,
obtained from $t$ by removing all symbols in $\Delta \setminus \Deltai$.
%\end{equation}
%
\noindent
\emph{Symbolic weighted parsing} is the problem,
given the above input,
to find $t \in \Delta^*$ %nested word $t \in \Delta$
minimizing \( d(s, t) \otimes A(t)\)
\wrt $\leq_\oplus$,
\ie s.t. %such that: %Hence, it is the problem of finding
%
\begin{equation}\label{eq:distance-lang}
d(s, t) \otimes A(t) = \displaystyle\bigoplus_{t' \in \Delta^*} d(s, t') \otimes A(t')
\end{equation}
%
Following the terminology of~\cite{Mohri03EDWA},
$\SW$-parsing is the problem of computing
the distance~\eqref{eq:distance-lang} between the input~$s$ and the output weighted language of~$A$,
and returning a witness $t$.
%
%\subsection{Computation}
%
\begin{proposition}
The problem of Symbolic Weighted  parsing
can be solved in PTIME in the size of the input \SWT $T$, \SWVPA $A$
and input word $s$,
and the computation time of the functions and operators of the label theory.
\end{proposition}
%
\begin{proof} (sketch)
We follow a \emph{Bar-Hillel} construction, for parsing by intersection.
%
\noindent
Let us first extend the \SWT $T$ over $\Sigma$, $\Deltai$
into a \SWT $T'$ over $\Sigma$ and $\Delta$
(and the same semiring and label theory $\Semiring$ and $\bar\Phi$),
such that for all $u \in \Sigma^*$, and $t \in {\Delta}^*$,
$T'(u, t) = T(u, t|_{\Deltai})$.
%
The transducer $T'$ simply skips every symbol
$b \in {\Delta} \setminus \Deltai$,
by the addition to $T$,
of new transitions of the form $\wei_{01}(q, \varepsilon, b, q')$.
%
\noindent
Then, using Corolary~\ref{cor:epsilon},
we construct from the input word $s \in \Sigma^*$ and $T'$
a \SWA $B_{s, T'}$,
such that for all $t \in \Delta^*$, $B_{s, T'}(t) = d(s, t)$.
%
%This automaton is such that for all $t \in \Delta^*$,
%\[
%   A_{T', s}\bigl(\lin(t)\bigr)
% = T'\bigl(s, \lin(t)\bigr)
% = T'\bigl(s, \lin(t)|_{\Deltai}\bigr)
% = d(s, t).
%\]
%
\noindent
Next, %we convert the input \SWTA $A$ over $\Delta$
%into a \SWVPA $A'$ over $\hat\Delta$, using Proposition~\ref{lem:SWTA}, and
we compute the \SWVPA $B_{s, T'} \otimes A$,
using Proposition~\ref{prop:SWVPA-product}.
%
\noindent
It remains to compute a best nested-word $t \in {\Delta}^*$
using the best-search procedure of Proposition~\ref{th:best-search}.
%and convert it into a best tree in $\T_\Delta$ in order to solve SW parsing
%for~$T$, $A$ and~$s$.
%\qed
\end{proof}
The $\SW$-parsing generalizes
the problem of searching the best derivation (AST) of a weighted CF-grammar $G$
that yields a given input word $w$.
The latter problem, sometimes called~\emph{weighted parsing},
(see \eg~\cite{Goodman99SemiringParsing}
 and~\cite{MorbitzVogler19weighted-parsing} for general weighted parsing frameworks)
corresponds to $\SW$-parsing in the case of finite alphabets,
a transducer $T$ computing the identity and some \SWVPA~$A$
obtained from $G$. %the weighted CF grammar.
%
%Indeed, it corresponds to the case where $T$
%accepts only the pairs $\<s, t>$ such that
%$s$ is the projection of $t$ on $\Deltai$.
%This can be done with a single state $q$ and
%with transition rules of the form:
%\begin{description}
%\item[] $\wei(q, \varepsilon, a, q) = \one$ for all $a \in \Deltac \cup \Deltar$,
%\item[] $\wei(q, a, a, q) = \one$ for all $a \in \Deltai$,
%\item[] $\wei(q, a, b, q) = \zero$ for all $a, b \in \Deltai$, $a \neq b$.
%\end{description}
%
Indeed, the \emph{depth-first} traversal of an AST $\tau$
yields a well-parenthesised word $\lin(\tau)$ over an alphabet
$\Delta = \Deltai \uplus \Deltac \uplus \Deltar$,
assuming \eg that $\Deltai$ contains the symbols labelling the leaves of $\tau$ (symbols of rank $0$),
and $\Deltac$ and $\Deltar$ contain respectively one left and right parenthesis
$\ccall{b}$ and $\creturn{b}$ for each symbol $b$ labelling inner nodes of $\tau$ (symbols of rank $>0$).
%
%With this representation, the projection $\lin(t)|_\Deltai$ is the
%sequence of leaves of $\tau$. %, enumerated in a \emph{dfs}-traversal.
%
We show in Appendix~\ref{sec:trees} how to construct 
%($\SW$) tree automaton~$A$ %(in particular the TA computing on the AST of a given CF-grammar)
a $\SWVPA$ $A$ such that $A(\lin(\tau))$ is the weight the AST $\tau$ of $G$.

>>>>

\noindent
In order to compute the above function $b_\bot : Q \times Q \to \Semiring$, 
we shall consider an auxiliary function $b_\top :  Q \times P \times Q \to \Phic$.
%
Intuitively, $b_\top(q, p, q')$ is a function of $\Phic$, 
mapping every $c \in \Deltac$ to 
the minimum weight of a computation of~$A$
starting in state $q$ with a non-empty stack 
$\gamma' = \<c, p> \, \gamma \in \Gamma^+$, 
and ending in state $q'$ with the same stack $\gamma'$,
such that moreover, 
the computation does not pop the pair $\<c, p>$ at the top of $\gamma'$
(\ie $\gamma'$ is left untouched during the computation).
However, the computation can read $\<c, p>$ at the top of $\gamma'$,
and can also push another pair $\< c', p'> \in \Gamma$ on $\gamma'$,
following the third case of 
in the definition \eqref{eq:SWVPA-weight} of $\weight_A$ (call symbol).
The pair $\< c', p'>$ can be pop later, during the computation from $q$ to $q'$,  
following the fifth case of \eqref{eq:SWVPA-weight} (return symbol). 
%However, it cannot apply one of the two last cases (return symbol and empty stack)
%when the current stack is $\gamma$.
%pop symbols in $\gamma$.
%
% Note that having a stack reduced to such a symbol makes impossible the application of the
% two last cases in the definition of $\weight_A$ (return symbol and empty stack).
% However, it is possible to apply the two first cases
% (internal symbol or call symbol, with a push on the top of $\top$).
%
Formally, in order to define $b_\top$, we consider 
a fresh stack symbol $\top \notin \Gamma$,   %which does not belong to $\Gamma$,
representing the above untouched stack, and let:
%
\begin{equation}\label{eq:btop}
  b_\top(q, p, q') : c \mapsto \bigoplus_{s\in \Delta^*}
  \textstyle
  \weight_A\bigl(\configup{q}{\< c, p> \stackup \top }, s, \configup{q'}{\<c, p> \stackup \top}\bigr)
\quad  
\mathrm{~for~all~} c \in \Deltac  
\end{equation}
%
By definition of $\weight_A$ in \eqref{eq:SWVPA-weight}, 
using the symbol $\top$ for the part of the stack below $\<c, p>$
(\ie the substack $\gamma$ in the above $\gamma' = \<c, p> \, \gamma$)
ensures that this part $\gamma'$ is not touched during the computation.
% there cannot be a pop of $\<c, p>$ followed by a push of $\<c, p>$
% because it is not possible to push over $\top$ (push read the top of stack).
%
This ensures in particular that the subword 
read during the computation is well parenthesized
(every symbol in $\Deltac$ has a matching symbol in $\Deltar$).

@@

\item[--] a \SWVPA $A$ over $\Delta$, $\Semiring$, and $\bar\Phi$,
defining a measure $A: \Delta^* \to \Semiring$,
%a \SWVPA $A$ over $\Delta$, and $\Semiring$, defining a series of nested words
%      $A : \Delta^* \to \Semiring$,
\item[--] an input word $s \in \Sigma^*$.
\end{description}
%
For all $u \in \Sigma^*$ and $t \in \Delta^*$, let
\(d(u, t) = T\bigl(u, t|_\Deltai \bigr)\),
where $t|_\Deltai \in \Deltai^*$ is the projection of $t$ onto~$\Deltai$,
obtained from $t$ by removing all symbols in $\Delta \setminus \Deltai$.
%\end{equation}
%
\noindent
\emph{Symbolic weighted parsing} is the problem,
given the above input,
to find $t \in \Delta^*$ %nested word $t \in \Delta$
minimizing \( d(s, t) \otimes A(t)\)
\wrt $\leq_\oplus$,
\ie s.t. %such that: %Hence, it is the problem of finding
%
\begin{equation}\label{eq:distance-lang}
d(s, t) \otimes A(t) = \displaystyle\bigoplus_{t' \in \Delta^*} d(s, t') \otimes A(t')
\end{equation}
%
Following the terminology of~\cite{Mohri03EDWA},
$\SW$-parsing is the problem of computing
the distance~\eqref{eq:distance-lang} between the input~$s$ and the output weighted language of~$A$,
and returning a witness $t$.

\begin{example}[Symbolic Weighted Parsing and the transcription problem]
Applied to the music transcription problem, the above formalism
is interpreted as follows:
\begin{enumerate}

  \item The \SWT $T$ evaluates a ``fitness measure''  that expresses
    a correspondance between a performance and its notation. See Example~\ref{ex:SWT}.
\item The \SWVPA $A$ expresses a cost related to the music notation. One possibility
 is to relate this cost to the structural complexity: given a set of equivalent
 representations, we aim at choosing the simpler one. For instance
 \includegraphics[scale=0.20]{pictures/score4.png} should be favored
 on \includegraphics[scale=0.20]{pictures/score5.png}, although both yield
 a same linearization. Costs  can be expressed
 in the grammar with weight associated to each production rule. See Example~\ref{rules-division}.

 \end{enumerate}
Therefore, the framework, applied to the transcription problem, allows to find an optimal
solution that considers both the fitness of the result to the input, and the complexity of the former.
\end{example}
%
%\subsection{Computation}
%
\begin{proposition}
The problem of Symbolic Weighted  parsing
can be solved in PTIME in the size of the input \SWT $T$, \SWVPA $A$
and input word $s$,
and the computation time of the functions and operators of the label theory.
\end{proposition}
%
\begin{proof} (sketch)
We follow a \emph{Bar-Hillel} construction, for parsing by intersection.
%
\noindent
Let us first extend the \SWT $T$ over $\Sigma$, $\Deltai$
into a \SWT $T'$ over $\Sigma$ and $\Delta$
(and the same semiring and label theory $\Semiring$ and $\bar\Phi$),
such that for all $u \in \Sigma^*$, and $t \in {\Delta}^*$,
$T'(u, t) = T(u, t|_{\Deltai})$.
%
The transducer $T'$ simply skips every symbol
$b \in {\Delta} \setminus \Deltai$,
by the addition to $T$,
of new transitions of the form $\wei_{01}(q, \varepsilon, b, q')$.
%
\noindent
Then, using Corolary~\ref{cor:epsilon},
we construct from the input word $s \in \Sigma^*$ and $T'$
a \SWA $B_{s, T'}$,
such that for all $t \in \Delta^*$, $B_{s, T'}(t) = d(s, t)$.
%
%This automaton is such that for all $t \in \Delta^*$,
%\[
%   A_{T', s}\bigl(\lin(t)\bigr)
% = T'\bigl(s, \lin(t)\bigr)
% = T'\bigl(s, \lin(t)|_{\Deltai}\bigr)
% = d(s, t).
%\]
%
\noindent
Next, %we convert the input \SWTA $A$ over $\Delta$
%into a \SWVPA $A'$ over $\hat\Delta$, using Proposition~\ref{lem:SWTA}, and
we compute the \SWVPA $B_{s, T'} \otimes A$,
using Proposition~\ref{prop:SWVPA-product}.
%
\noindent
It remains to compute a best nested-word $t \in {\Delta}^*$
using the best-search procedure of Proposition~\ref{th:best-search}.
%and convert it into a best tree in $\T_\Delta$ in order to solve SW parsing
%for~$T$, $A$ and~$s$.
%\qed
\end{proof}
The $\SW$-parsing generalizes
the problem of searching the best derivation (AST) of a weighted CF-grammar $G$
that yields a given input word $w$.
The latter problem, sometimes called~\emph{weighted parsing},
(see \eg~\cite{Goodman99SemiringParsing}
 and~\cite{MorbitzVogler19weighted-parsing} for general weighted parsing frameworks)
corresponds to $\SW$-parsing in the case of finite alphabets,
a transducer $T$ computing the identity and some \SWVPA~$A$
obtained from $G$. %the weighted CF grammar.
%
%Indeed, it corresponds to the case where $T$
%accepts only the pairs $\<s, t>$ such that
%$s$ is the projection of $t$ on $\Deltai$.
%This can be done with a single state $q$ and
%with transition rules of the form:
%\begin{description}
%\item[] $\wei(q, \varepsilon, a, q) = \one$ for all $a \in \Deltac \cup \Deltar$,
%\item[] $\wei(q, a, a, q) = \one$ for all $a \in \Deltai$,
%\item[] $\wei(q, a, b, q) = \zero$ for all $a, b \in \Deltai$, $a \neq b$.
%\end{description}
%
Indeed, the \emph{depth-first} traversal of an AST $\tau$
yields a well-parenthesised word $\lin(\tau)$ over an alphabet
$\Delta = \Deltai \uplus \Deltac \uplus \Deltar$,
assuming \eg that $\Deltai$ contains the symbols labelling the leaves of $\tau$ (symbols of rank $0$),
and $\Deltac$ and $\Deltar$ contain respectively one left and right parenthesis
$\ccall{b}$ and $\creturn{b}$ for each symbol $b$ labelling inner nodes of $\tau$ (symbols of rank $>0$).
%
%With this representation, the projection $\lin(t)|_\Deltai$ is the
%sequence of leaves of $\tau$. %, enumerated in a \emph{dfs}-traversal.
%
We show in Appendix~\ref{sec:trees} how to construct
%($\SW$) tree automaton~$A$ %(in particular the TA computing on the AST of a given CF-grammar)
a $\SWVPA$ $A$ such that $A(\lin(\tau))$ is the weight the AST $\tau$ of $G$.
